Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2013-09-03 paper.Rnw
Fixes for Mallar's recent edits

% styling
\newcommand{\subsubsubsection}[1]{\paragraph{#1}}
\renewcommand\Affilfont{\itshape\small}  %authblk

% shortcuts 
\newcommand{\mb}{MAGeT-Brain }
\newcommand{\ants}{ANTS }
\newcommand{\animal}{ANIMAL }
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T }
\newcommand{\FEPdataset}{SZ-FEP }
\newcommand{\gs}{Gold Standard }
\newcommand{\fsl}{FSL }
\newcommand{\freesurfer}{FreeSurfer }
\newcommand{\maper}{MAPER }

% title and authors
\title{Bootstrapping Multi-atlas Hippocampal Subfield Segmentation by using
Multiple Automatically Generated Templates}
\author[1]{Jon Pipitone}
\author[1]{Matt T. Park}
\author[1]{Julie Winterburn}
\author[1,8]{Tristram A. Lett}
\author[2,3]{Jason P. Lerch}
\author[4,5]{Jens C. Pruessner}
\author[4]{Martin Lepage}
\author[1,6,8]{Aristotle N. Voineskos}
\author[1,6,7]{M. Mallar Chakravarty}
\author[ ]{the Alzheimer's Disease Neuroimaging Initiative\footnote{Data
used in preparation of this article were obtained from the Alzheimer's Disease
Neuroimaging Initiative (ADNI) database (adni.loni.ucla.edu). As such, the
investigators within the ADNI contributed to the design and implementation of
ADNI and/or provided data but did not participate in analysis or writing of this
report. A complete listing of ADNI investigators can be found at:
http://adni.loni.ucla.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}
\affil[1]{Kimel Family Translational Imaging-Genetics Lab, Centre for Addiction and
Mental Health, Toronto, ON, Canada}
\affil[2]{Neurosciences and Mental Health, Hospital for Sick Children, Toronto,
ON,Canada}
\affil[3]{Department of Medical Biophysics, University of Toronto, Toronto, ON,
Canada}
\affil[4]{Douglas Mental Health University Institute, Verdun, QC, Canada}
\affil[5]{Department of Psychiatry, McGill University, Montreal, QC, Canada}
\affil[6]{Department of Psychiatry, University of Toronto, Toronto, ON, Canada}
\affil[7]{Institute of Biomedical Engineering, University of Toronto, Toronto,
ON, Canada}
\affil[8]{Institute of Medical Science, University of Toronto, Toronto, ON, Canada}

@@

\textbf{Introduction:} Neuroimaging research often relies on automated anatomical
segmentations of MR images of the brain. Current multi-atlas based approaches
provide accurate segmentations of brain images by propagating manually derived
segmentations of specific neuroanatomical structures to unlabelled data. These
approaches often rely on a large number of such manually segmented atlases (e.g.
30-80) that take significant time and expertise to produce. We present an
algorithm, \mb, for the automatic segmentation of the hippocampus that minimizes
the number of atlases needed while still achieving similar accuracy to
multi-atlas approaches.\\
\textbf{Method:} \mb works by propagating atlas segmentations to a template library,
formed from a subset of target images, via transformations estimated by non-linear
image registration. The resultant segmentations are then propagated to each
target image and fused using a label fusion method. We conduct a 10-fold Monte
Carlo cross-validation of whole hippocampal segmentation on a pool of 69 ADNI
subjects over a range of parameter settings, and a leave-one-out
cross-validation (LOOCV) of hippocampal subfield segmentation using five
high-resolution atlases. Two final experiments assess \mb when applied to
Schizophrenia and Alzheimer's disease populations, and \mb segmentations are
compared with existing automated methods (\fsl, \freesurfer, \maper) and biases are
explored.\\
\textbf{Results:} Using nine atlases and 19 template images, \mb achieves a
mean Dice's Similarity Coefficient (DSC) of 0.84 over 10-folds of Monte Carlo
cross-validation on 69 subjects relative to semi-automated \gs segmentations,
and shows significantly lower variability in DSC than multi-atlas segmentation.
In a LOOCV, \mb reproduces the volumes of the CA1, CA4/DG, and SR/SL/SM
hippocampal subregions with a percent error in volume that is at or lower than
that produced by image resampling. \mb produces hippocampal volumes in a
Schizophrenia patient population that are highly correlated with expert manual
segmentation volumes (Pearson $r = 0.877, t = 16.244, p < 0.001$). Compared to
\fsl and \freesurfer, \mb shows much smaller fixed volume bias (within $250
mm^3$ on average) to \gs segmentations, as well as a conservative, rather than
exaggerated, proportional volume bias.\\
\textbf{Conclusion:} We demonstrate that \mb produces accurate hippocampal
segmentations using only 5 atlases over different hippocampal definitions,
disease populations, and acquisition types, as well as showing that accurate
identification of the hippocampal subfields is possible.
\end{abstract}

@@

The hippocampus is a brain structure situated in the medial temporal
lobe, and has long been associated with learning and memory
\citep{DenHeijer2012,Scoville2000}. The hippocampus is of interest to
neuroscientists because it is implicated in many forms of brain dysfunction,
including Alzheimer's disease \citep{Sabuncu2011} and schizophrenia
\citep{Narr2004,Karnik-Henry2012}. In neuroimaging studies, structural magnetic
resonance images (MRI) are often used for the volumetric assessment of the
hippocampus.  As such, accurate segmentation of the hippocampus and its
subregions in MRI is a necessary first step to better understand the
inter-individual variability of subject neuroanatomy.

Typically, the gold standard for neuroanatomical segmentation is manual
delineation by an expert human rater.  However, with the availability of
increasingly large MRI datasets, the time and expertise required for manual
segmentation becomes prohibitive
\citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007}. Further, there is
significant variation between segmentation protocols regarding specific
anatomical boundaries of the hippocampus \citep{Geuze2004} and this has led to
efforts to create an unified hippocampal segmentation protocol \citep{Jack2011}.

@@

To the best of our knowledge there are only two automated segmentation methods
that attempt to define the hippocampus and its subfields using standard
T1-weighted data. The first is included with the \freesurfer package
\citep{VanLeemput2009}.  This work is limited as it omits the tail of the
hippocampus and the segmentation protocol has yet to be fully validated.
Nonetheless, it demonstrates that the applicability of hippocampal subfield
segmentation using data from 10 subjects. In the second method,
\citet{Yushkevich2009} hippocampal subfields were acquired and labelled on
high-resolution MRI data from post-mortem medial temporal lobe samples. Using
nonlinear registration guided by manually derived hippocampus masks and specific
landmarks, the authors demonstrated accurate parcellation of hippocampal
subfields in unlabelled MRI volumes.

In this paper we undertook a thorough validation of the \mb algorithm for
segmentation of the hippocampus. First, we addressed the feasibility and
accuracy of whole hippocampus segmentation with a limited number of input
atlases by performing a multi-fold experiment using a subset of the Alzheimer's
Disease Neuroimaging Initiative (ADNI) dataset over a range of parameters.  We
then performed a leave-one-out validation to determine if hippocampal subfields
can be accurately identified using our multi-atlas framework.  To ensure that we
have not over-fit our parameters to the aging or neurodegenerative brain, we
also applied our segmentations to a dataset of individuals suffering from first
episode psychosis.  Finally, validated our algorithm using all of the data
available in the \adnidataset sample and compare our segmentations to the other
segmentations available through the ADNI informatics portal.

@@

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset used was the "\adnidataset" standardized dataset
available from ADNI \footnote{
\url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly pre-processed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008}.  Representative 1.5T imaging parameters were TR =
2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8{\circ}$, field of view = 240 x
240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding a voxel dimensions of $1.25mm \times 1.25mm \times 1.2mm$.

@@

In addition to the high-resolution scans distributed as part of the Winterburn
atlases, we also obtained separate 3T T1 BRAVO acquisitions (0.9mm-isotropic
voxels) of four of the five Winterburn atlas subjects.

@@

In part 2a of this experiment, we have modified the LOOCV experimental
design in order to compare \mb segmentation accuracy with the accuracy of image
resampling error. The five $0.3mm$-isotropic voxel Winterburn atlas images and
labels are used as the atlas library, and LOOCV is conducted using the five Winterburn
atlas images subsampled (using trilinear interpolation) to
0.9mm-isotropic voxel resolution (referred to as the {\em Subsampled} dataset)
as input subjects. Subsampling the target images allows us to assess \mb in a
typical segmentation scenario (high-resolution atlases and lower-resolution
subjects).  The template library is composed of all five subject images, plus an
additional set of 3T T1 images (0.9mm-isotropic voxels) of healthy subjects
acquired separately (Table \ref{tab:WAval-healthy-demographics}). The optimal
size of template library, registration method, and label fusion method found in
Experiment 1 are used.  As with traditional LOOCV, each subject image is
segmented by \mb with that subject's image excluded from the atlas library.  In
part 2b of this experiment, we carry out a similar validation in which the
subject images are separately acquired T1 BRAVO images of four of the five
Winterburn atlas subjects (referred to as the {\em T1 BRAVO} dataset).  Figure
\ref{fig:LOOCV-schematic} illustrates the experimental set up of both parts of
this experiment.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Subfield Validation Template library demographics (healthy controls)
%                                     
<<WAval-healthy-demographics, echo=F, results="asis",cache=FALSE>>=
master = read.csv('data/DT_study.csv')
tab <- summary( Diagnosis ~ Age + Sex + Education + Handedness, data = master,
    method="reverse", na.rm = TRUE, test=FALSE)
caption = 
  "\\textbf{Subfield validation healthy control demographics.}
  CN     - Cognitively Normal. 
  LMCI   - Late-onset Mild Cognitive Impairment. 
  AD     - Alzheimer's Disease.  
  Hisp   - Hispanic. 
  CDR-SB - Clinical Dementia Rating-Sum of Boxes.  
  ADAS   - Alzheimer's Disease Assessment Scale.  
  MMSE   - Mini-Mental State Examination."
latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption=caption, caption.loc = c("top"),
      title="tab:WAval-healthy-demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Scanning of 81 subjects was carried out at the Montreal Neurological Institute
on a 1.5-T Siemens whole body MRI system.  Structural T1 volumes were acquired
for each participant using a three-dimensional (3D) gradient echo pulse sequence
with sagittal volume excitation (repetition time=22ms, echo time=9.2ms, flip
angle=$30{\circ}$, 180 1mm contiguous sagittal slices). The rectangular
field-of-view for the images was 256mm (SI)$\times$204mm (AP).  Subject
demographics are shown in Table \ref{tab:SZFEP-Demographics}.

@@

\paragraph{Evaluation method}
The Pruessner and Winterburn hippocampal segmentation protocols differ slightly in
the neuroanatomical features that are delineated \citep{Winterburn2013}. This
difference poses a problem for evaluation by measuring overlap.  That is, since
different protocols will necessarily produce segmentations that do not perfectly
overlap, the degree of overlap cannot be solely used to compare segmentation
methods using different protocols. In place of an overlap metric, we can assess
the degree of correlation in average bilateral hippocampal volume of the
subjects produced by each method. Specifically, a Pearson correlation is used.

@@

In addition to an improvement in accuracy over multi-atlas-based segmentation,
\mb also shows a decrease in the variability of segmentation accuracy 
(Figure \ref{ADNI1-xval-variability}).  The size of template library 
necessary to reach a significant ($p < 0.05$) decrease in variance and standard 
deviation grows with the size of atlas library used.  A template library of 19
images is sufficient to show significant decrease in variance and standard 
deviation for 3-7 atlases.

@@

\begin{sidewaysfigure}
    \begin{centering}
      \includegraphics[width=\linewidth]{figure/winterburn-atlas-montage/figure.pdf}
    \end{centering}
    \caption{Detailed subfield segmentation results of a single subject comparing the
    original Winterburn atlas subfield segmentation (at 0.3mm-isotropic voxel
    resolution), the Winterburn segmentation resampled (at 0.9mm-isotropic
    voxel resolution), the \mb segmentation of the subsampled atlas image, and
    the \mb segmentation of a separate T1 BRAVO acquisition of the subject.}
    \label{fig:subfield-montage}
\end{sidewaysfigure}

@@

On that note, it is worth observing that we found the SNT segmentations to be
poor in general. As previously noted in Figure \ref{fig:ADNI-segmentations}, an
expert manual rater (JW) identified many inconsistencies: occurrences of over-
and under-estimation, as well as misalignments of the entire segmentation
volume.  Therefore, the use of the SNT segmentations as gold standards in
Experiments 1 and 4 presents the possibility that the DSC scores and
correlations we obtained are lower than what would be found with a more accurate
and consistent gold standard.

>>>>

% styling
\newcommand{\subsubsubsection}[1]{\paragraph{#1}}
\renewcommand\Affilfont{\itshape\small}  %authblk
\linespread{1.3} % one and a half line spacing

% shortcuts 
\newcommand{\mb}{MAGeT-Brain}
\newcommand{\ants}{ANTS}
\newcommand{\animal}{ANIMAL}
\newcommand{\adnidataset}{ADNI1:Complete 1Yr 1.5T}
\newcommand{\FEPdataset}{SZ-FEP}
\newcommand{\gs}{Gold Standard}
\newcommand{\fsl}{FSL}
\newcommand{\freesurfer}{FreeSurfer}
\newcommand{\maper}{MAPER}

% title and authors
%\title{Bootstrapping Multi-atlas Hippocampal Subfield Segmentation by using
%Multiple Automatically Generated Templates}
\title{Bootstrapping Multi-atlas Segmentation Using Multiple Automatically
Generated Templates for the Segmentation of the Whole Hippocampus and
Subfields}
\author[1]{Jon Pipitone}
\author[1]{Matt T. Park}
\author[1]{Julie Winterburn}
\author[1,9]{Tristram A. Lett}
\author[2,3]{Jason P. Lerch}
\author[4,5]{Jens C. Pruessner}
\author[4]{Martin Lepage}
\author[1,6,9]{Aristotle N. Voineskos}
\author[1,6,8]{M. Mallar Chakravarty}
\author[ ]{the Alzheimer's Disease Neuroimaging Initiative\footnote{Data
used in preparation of this article were obtained from the Alzheimer's Disease
Neuroimaging Initiative (ADNI) database (adni.loni.ucla.edu). As such, the
investigators within the ADNI contributed to the design and implementation of
ADNI and/or provided data but did not participate in analysis or writing of this
report. A complete listing of ADNI investigators can be found at:
http://adni.loni.ucla.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}
\affil[1]{Kimel Family Translational Imaging-Genetics Lab, Centre for Addiction and
Mental Health, Toronto, ON, Canada}
\affil[2]{Neurosciences and Mental Health Laboratory, Hospital for Sick
Children, Toronto, ON, Canada}
\affil[3]{Department of Medical Biophysics, University of Toronto, Toronto, ON,
Canada}
\affil[4]{Douglas Mental Health University Institute, Verdun, QC, Canada}
\affil[5]{Department of Psychiatry, McGill University, Montreal, QC, Canada}
\affil[6]{Department of Psychiatry, University of Toronto, Toronto, ON, Canada}
\affil[7]{Institute of Biomaterials and Biomedical Engineering, University of
Toronto, Toronto, ON, Canada}
\affil[8]{Rotman Research Institute, Baycrest, Toronto, ON, Canada}
\affil[9]{Institute of Medical Science, University of Toronto, Toronto, ON, Canada}

@@

\textbf{Introduction:} 
Advances in image segmentation of magnetic resonance images (MRI) have
demonstrated that multi-atlas approaches improve segmentation accuracy and
precision over regular atlas-based approaches.  These approaches often rely on
a large number of such manually segmented atlases (e.g.  30-80) that take
significant time and expertise to produce. We present an algorithm, \mb (\mb
({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates), for the automatic segmentation of the hippocampus that minimizes
the number of atlases needed while still achieving similar accuracy to
multi-atlas approaches. Thus, our method acts as an accurate multi-atlas
approach when using special, hard-to-define atlases that are laborious to
construct. \\
\textbf{Method:} \mb works by propagating atlas segmentations to a template library,
formed from a subset of target images, via transformations estimated by non-linear
image registration. The resultant segmentations are then propagated to each
target image and fused using a label fusion method. We conduct a 10-fold Monte
Carlo cross-validation of whole hippocampal segmentation on a pool of 69 ADNI
subjects over a range of parameter settings, and a leave-one-out
cross-validation (LOOCV) of hippocampal subfield segmentation using five
high-resolution atlases. Two final experiments assess \mb when applied to
first episode psychosis and Alzheimer's disease populations, and \mb
segmentations are compared with existing automated methods (\fsl, \freesurfer,
\maper) and biases are explored.\\
\textbf{Results:} Using nine atlases and 19 template images, \mb achieves a
mean Dice's Similarity Coefficient (DSC) of 0.84 over 10-folds of Monte Carlo
cross-validation on 69 subjects relative to semi-automated \gs segmentations,
and shows significantly lower variability in DSC than multi-atlas segmentation.
In a LOOCV, \mb reproduces the volumes of the cornu ammonis (CA) 1; CA4/dentate
gyrus (DG); and strata radiatum (SR), strata lacunosum (SL), and strata
moleculare (SM) hippocampal subregions with a percent error in volume that is
at or lower than that produced by image resampling. \mb produces hippocampal
volumes in a first episode psychosis patient population that are highly
correlated with expert manual segmentation volumes (Pearson $r = 0.877, t =
16.244, p < 0.001$). Compared to \fsl and \freesurfer, \mb shows much smaller
fixed volume bias (within $250 mm^3$ on average) to \gs segmentations available
from ADNI, as well as a conservative, rather than exaggerated, proportional
volume bias.\\ \textbf{Conclusion:} We demonstrate that \mb produces accurate
hippocampal segmentations using only 5 atlases over different hippocampal
definitions, disease populations, and acquisition types, as well as showing
that accurate identification of the hippocampal subfields is possible.
\end{abstract}

@@

The hippocampus is a brain structure situated in the medial temporal
lobe, and has long been associated with learning and memory
\citep{DenHeijer2012,Scoville2000}. The hippocampus is of interest to
clinical neuroscientists because it is implicated in many forms of brain
dysfunction, including Alzheimer's disease \citep{Sabuncu2011} and
schizophrenia \citep{Narr2004,Karnik-Henry2012}. In neuroimaging studies,
structural magnetic resonance images (MRI) are often used for the volumetric
assessment of the hippocampus.  As such, accurate segmentation of the
hippocampus and its subregions in MRI is a necessary first step to better
understand the inter-individual variability of subject neuroanatomy.

The gold standard for neuroanatomical segmentation is manual delineation by an
expert human rater.  However, with the availability of increasingly large MRI
datasets, the time and expertise required for manual segmentation becomes
prohibitive \citep{Mazziotta1995,Mazziotta2001,Mazziotta,Pausova2007}. This
effort is complicated by the fact that there is significant variation between
segmentation protocols with respect to specific anatomical boundaries of the
hippocampus \citep{Geuze2004} and this has led to efforts to create an unified
hippocampal segmentation protocol \citep{Jack2011}.  In addition, there is
controversy over the appropriate manual segmentation protocol to use in a
particular imaging study \citep{Nestor2012}.  Thus, a segmentation algorithm
that can easily adapt to different manual segmentation definitions would be of
significant benefit to the neuroimaging community.

@@

Also of interest to this manuscript are methods that attempt to define
hippocampal subfields using standard T1-weighted data.  To the best of our
knowledge, there are only two automated segmentation algorithms that attempt
this problem. The first is included with the \freesurfer package
\citep{VanLeemput2009}. This work is limited as it omits the tail of the
hippocampus and the segmentation protocol has yet to be fully validated.
Nonetheless, it demonstrates that the applicability of hippocampal subfield
segmentation using data from 10 subjects. In the second method,
\citet{Yushkevich2009} hippocampal subfields were acquired and labelled on
high-resolution MRI data from post-mortem medial temporal lobe samples. Using
nonlinear registration guided by manually derived hippocampus masks and
specific landmarks, the authors demonstrated accurate parcellation of
hippocampal subfields in unlabelled MRI volumes.

In this paper we have describe a thorough validation of the \mb algorithm for
segmentation of the hippocampus and its subfields. First, we addressed the feasibility and
accuracy of whole hippocampus segmentation with a limited number of input
atlases \citep{MallarChakravarty2012} by performing a multi-fold experiment
using a subset of the Alzheimer's Disease Neuroimaging Initiative (ADNI)
dataset over a range of parameters.  We then performed a leave-one-out
validation to determine if hippocampal subfields can be accurately identified
using our multi-atlas framework.  To ensure that we have not over-fit our
parameters to the aging or neurodegenerative brain, we also applied our
segmentations to a dataset of individuals suffering from first episode
psychosis.  Finally, validated our algorithm using all of the data available in
the \adnidataset sample and compare our segmentations to the other
segmentations available through the ADNI informatics portal.

@@

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset used was the "\adnidataset" standardized dataset
available from ADNI \footnote{
\url{http://adni.loni.ucla.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly pre-processed images
which have been designated to be the "best" after quality control.  All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citep{Jack2008}.  Representative 1.5T imaging parameters were TR =
2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8^{\circ}$, field of view = 240 x
240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding a voxel dimensions of $1.25mm \times 1.25mm \times 1.2mm$.

@@

In addition to the high-resolution scans distributed as part of the Winterburn
atlases, we also obtained separate 3T T1 BRAVO acquisitions (0.9mm-isotropic
voxels) of four of the five Winterburn atlas subjects.  Images were acquired on
a 3T GE Discovery MR 750 system (General Electric, Milwaukee, WI) using an
8-channel head coil with the enhanced fast gradient recalled echo 3-dimensional
acquisition protocol, EFGRE-BRAVO, with the following parameters: $TE/TR/TI =
3.0ms/6.7ms/650ms$, flip angle $=8^\circ$, $FOV=15.8cm$, slice
thickness$=0.9mm$, 176 in-plane steps for an approximate isotropic resolution
of 0.9mm dimension voxels.

@@

In part 2a of this experiment, we have modified the LOOCV experimental
design in order to compare \mb segmentation accuracy with the accuracy of image
resampling error. The five $0.3mm$-isotropic voxel Winterburn atlas images and
labels are used as the atlas library, and LOOCV is conducted using the five Winterburn
atlas images subsampled (using trilinear interpolation) to
0.9mm-isotropic voxel resolution (referred to as the {\em Subsampled} dataset)
as input subjects. Subsampling the target images allows us to assess \mb in a
typical segmentation scenario (high-resolution atlases and lower-resolution
subjects).  The template library is composed of all five subject images, plus
an additional set of 16 of healthy subjects 3T T1 images acquired separately
(Table \ref{tab:WAval-healthy-demographics}).  Images were acquired on a 3T GE
Discovery MR 750 system (General Electric, Milwaukee, WI) using an 8-channel
head coil with the enhanced fast gradient recalled echo 3-dimensional
acquisition protocol, EFGRE-BRAVO, with the following parameters: $TE/TR/TI =
3.0ms/6.7ms/650ms$, flip angle=$8^\circ$ , $FOV = 15.3cm$, slice
thickness$=0.9mm$, 170 in-plane steps for an approximate isotropic resolution
of 0.9mm dimension voxels.

The optimal size of template library, registration method, and label fusion
method found in Experiment 1 are used.  As with traditional LOOCV, each subject
image is segmented by \mb with that subject's image excluded from the atlas
library.  In part 2b of this experiment, we carry out a similar validation in
which the subject images are separately acquired T1 BRAVO images of four of the
five Winterburn atlas subjects (referred to as the {\em T1 BRAVO} dataset).
Figure \ref{fig:LOOCV-schematic} illustrates the experimental set up of both
parts of this experiment.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Subfield Validation Template library demographics (healthy controls)
%                                     
<<WAval-healthy-demographics, echo=F, results="asis",cache=FALSE>>=
master = read.csv('data/DT_study.csv')
tab <- summary( Diagnosis ~ Age + Sex + Education + Handedness, data = master,
    method="reverse", na.rm = TRUE, test=FALSE)
caption = 
  "\\textbf{Demographics for the hippocampal subfield cross-validation healthy
control subject sample.} Education is shown in years."
latex(tab, file="", size="scriptsize",landscape=FALSE, ctable=FALSE,
      caption=caption, caption.loc = c("top"),
      title="tab:WAval-healthy-demographics")
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Scanning of 81 subjects was carried out at the Montreal Neurological Institute
on a 1.5-T Siemens whole body MRI system.  Structural T1 volumes were acquired
for each participant using a three-dimensional (3D) gradient echo pulse sequence
with sagittal volume excitation (repetition time=22ms, echo time=9.2ms, flip
angle=$30^{\circ}$, 180 1mm contiguous sagittal slices). The rectangular
field-of-view for the images was 256mm (SI)$\times$204mm (AP).  Subject
demographics are shown in Table \ref{tab:SZFEP-Demographics}.

@@

\paragraph{Evaluation method}
The Pruessner and Winterburn hippocampal segmentation protocols differ slightly in
the neuroanatomical features that are delineated \citep{Winterburn2013}. This
difference poses a problem for evaluation by measuring overlap.  That is, since
different protocols will necessarily produce segmentations that do not perfectly
overlap, the degree of overlap cannot be solely used to compare segmentation
methods using different protocols. In place of an overlap metric, we can assess
the degree of (Pearson) correlation in average bilateral hippocampal volume of the
subjects produced by each method.

@@

In addition to an improvement in accuracy over multi-atlas-based segmentation,
\mb also shows a decrease in the variability of segmentation accuracy 
(Figure \ref{ADNI1-xval-variability}).  The size of template library 
necessary to reach a decrease ($p < 0.05$) in variance and standard deviation
grows with the size of atlas library used.  A template library of 19 images is
sufficient to show significant decrease in variance and standard deviation for
3-7 atlases.

@@

\begin{sidewaysfigure}
    \begin{centering}
      \includegraphics[width=\linewidth]{figure/winterburn-atlas-montage/figure.pdf}
    \end{centering}
    \caption{Detailed subfield segmentation results for a single subject. In
    the upper left corner is the original high-resolution Winterburn atlas manual
    subfield segmentation; in the upper right corner is the Winterburn atlas
    segmentation resampled from 0.3mm- to 0.9mm-isotropic voxels; in the lower left
    corner is the \mb segmentation of the resampled Winterburn atlas images; in the
    lower right corner is the \mb segmentation of a separately acquired T1 BRAVO
    image of the same subject. In each segmentation, slices from the left
    hemisphere are shown in Talairach-like ICBM152 space: the first row shows axial
    slices from inferior to superior; the second row shows sagittal slices from
    lateral to medial; the third row shows coronal slices from anterior to
    posterior.  \label{fig:subfield-montage}}
\end{sidewaysfigure}

@@

On that note, one author (JW), an expert manual rater, identified regular
inconsistencies in the SNT segmentations: occurrences of over- and
under-estimation, as well as misalignments of the entire segmentation volume
(Figure \ref{fig:ADNI-segmentations}). Although the SNT segmentations are used
as gold standards for validation in many other studies (Table
\ref{tab:other-methods}) providing a benchmark for comparison, their use in
Experiments 1 and 4 presents the possibility that the DSC scores and
correlations we obtained are lower than what would be found with a more
accurate and consistent gold standard.
